Higher-Order unification for free!

### 2024-04-29 paper/main.tex
abstract no cite for copy paste

>>>>

\begin{comment}
Specifying and implementing a proof system from scratch requires significant
effort. Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level meta languages to facilitate this task in two ways: 1)
variable binding and substitution are for free when meta language binders
represent object logic ones; 2) proof construction, and
proof search, are greatly simplified by leveraging the unification procedure
provided by the meta language. Notable examples of meta languages are Elf,
Twelf, $\lambda$Prolog, Beluga, Abella and Isabelle which have been used to
implement or specify many formal systems such as First Order Logic, Set Theory,
Higher Order Logic, and the Calculus of Constructions.

The object logic we are interested in is Coq's type theory. We aim to develop a
higher-order unification-based proof search procedure using the meta language
Elpi, a dialect of $\lambda$Prolog. Elpi's equational theory includes
$\eta\beta$-equivalence and features a higher-order unification procedure \Ue
for the pattern fragment. Elpi offers an encoding of Coq terms that is suitable
for meta programming but that restricts \Ue to first-order unification problems
only. We refer to this basic encoding as \Fo.

In this paper we propose a more well-behaved encoding called \Ho, and show how
to translate unification problems from \Fo{} to corresponding ones in \Ho.
Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
that honours $\eta\beta$-equivalence (for Coq functions), addresses problems
within the pattern fragment, and allows for the use of heuristics to deal with
problems outside the pattern fragment. Moreover, as \Uo{} delegates most of the
work to \Ue, it can be used to efficiently simulate a logic program in \Fo{} by
taking advantage of unification-related optimizations of the meta language, such
as clause indexing.
\end{comment}
